---
title: "Starting with with the dplyr, ggplot2 packages, and the pipe operator |>"
author: "Han Olff"
date: "2024-08-29"
format: html
editor: 
  markdown: 
    wrap: 72
editor_options:
  chunck_output_type: console
  chunk_output_type: console
---

# Working with the dplyr and ggplot2 packages

Load the required libraries :

```{r message = FALSE, warning = FALSE}
# restore your library (keep at the start of every script)
renv::restore()

# clear the R environment
rm(list = ls())

# load the required packages
library(tidyverse)  # this loads a group of related packages, including readr, dplyr and ggplot2
```

We will work with an online database of the Schiermonnikoog transect
study, that you will collect additional data for next week in the field.
In this example, we work with measurements of cockles (a bivalve
mollusc) on their width and length. From the database Epibenthos, we
will work with the table
[FactCockles](https://docs.google.com/spreadsheets/d/1E1vUWAHhse7fhjBf94Kiog3Rsj7IkMuFtP-BTQX2oI8/edit?gid=1538766002#gid=1538766002).
See the documentation of the different variables in the table
[MetVariables](https://docs.google.com/spreadsheets/d/1E1vUWAHhse7fhjBf94Kiog3Rsj7IkMuFtP-BTQX2oI8/edit?gid=1290622213#gid=1290622213)

### Read the datafile from the Google sheets database

To read the data in R, you first need to know the published csv link of
the FactCockles table. In the database, you can find this link in the
table MetTables. It is produced in Google Sheets throught the menu
File/Share/Publish to web and then selecting the table and output as
csv.

Read the FactCockleSize table with read_csv, which reads it as tibble
(formatted dataframe)

```{r}
FactCockleSize <- readr::read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSpormjGiM4uWMTIZYm9Upt5j1Ige_Wu9CrhGGdvjXwxmHP2R6S7JDyEmiGTn3fXihl5s5yBn_sdo8h/pub?gid=1538766002&single=true&output=csv")

print(FactCockleSize)
names(FactCockleSize)
nrow(FactCockleSize)

```

Plot with ggplot the relation between cockle thickness (thickness_mm, as
x) and length (length_mm), showing each year with a different color, and
add a regression line through all the years

```{r}
ggplot2::ggplot(data=FactCockleSize,#what datatable to use
                mapping=aes(x=length_mm,y=thickness_mm)) + #what is on the axes
  geom_point()
#there is a big outlier, which is likely a data entry mistake, remove it (only when you have good reasons to do so, when biologically impossible).one cockle is 1500 mm thick, not possible.
FactCockleSize|> dplyr::filter(thickness_mm<500) #to remove the mistake

FactCockleSize <- readr::read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSpormjGiM4uWMTIZYm9Upt5j1Ige_Wu9CrhGGdvjXwxmHP2R6S7JDyEmiGTn3fXihl5s5yBn_sdo8h/pub?gid=1538766002&single=true&output=csv")|> dplyr::filter(thickness_mm<500) #in een keer goede data set zonder outlier

#then you can plot again. 

#better method: when you indentified the ID of the mistake:
FactCockleSize <- readr::read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSpormjGiM4uWMTIZYm9Upt5j1Ige_Wu9CrhGGdvjXwxmHP2R6S7JDyEmiGTn3fXihl5s5yBn_sdo8h/pub?gid=1538766002&single=true&output=csv")|> dplyr::filter(CockleObs_ID!=1531) #and then plot

#but there is another outlier: 3mm length, 15 mm thick
FactCockleSize|> dplyr::filter(thickness_mm>10 &length_mm<5) #gives the ID 469, remove this one as well
FactCockleSize <- readr::read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSpormjGiM4uWMTIZYm9Upt5j1Ige_Wu9CrhGGdvjXwxmHP2R6S7JDyEmiGTn3fXihl5s5yBn_sdo8h/pub?gid=1538766002&single=true&output=csv")|> dplyr::filter(CockleObs_ID!=1531, CockleObs_ID!=469)|> dplyr::mutate(year=factor(year))
  
print(FactCockleSize)
```

dplyr:
filter: select particular rows
select: select particular variables columns
mutate: transform calculate a variable
group_by: make group
summary: calculate
Further explore the plot with a regression line through all the data

```{r}
#plot data with regression line
ggplot2::ggplot(data=FactCockleSize,
                mapping=aes(x=length_mm,y=thickness_mm)) + 
   geom_point()+ geom_smooth(method = "lm")

#calculate the linear regression
model_lm<-lm(thickness_mm~length_mm, data = FactCockleSize)#(y~x)
summary(model_lm)
#equation of the regression line :thickness_mm= 0.70*length_mm - 0.35
# y=b0+b1x

#so, a cockle of length 20 mm has a predicted width of:
0.70*20 - 0.35 #=13.65 mm
#this regression is significant, because R2=0.90, p<0.001
```

Make same plot but showing a separate regression line per year

```{r}
# color the points by year, but plot one regression line
ggplot2::ggplot(data=FactCockleSize,
                mapping=aes(x=length_mm,y=thickness_mm)) + 
  geom_point(mapping =aes(col=year))+
  geom_smooth(method = "lm")
```

Make a panel plot where with each year is shown as a separate graph

```{r}
ggplot2::ggplot(data=FactCockleSize,
                mapping=aes(x=length_mm,y=thickness_mm)) + 
  geom_point(mapping =aes(col=year))+
  geom_smooth(method = "lm")+
  facet_wrap(~year)
#but there are years without data:
#taka dataset out of mapping, and pipe

drop_na(FactCockleSize)|>
ggplot2::ggplot(mapping=aes(x=length_mm,y=thickness_mm)) + 
  geom_point(mapping =aes(col=year),size=2)+
  xlab("Cockle length (mm)")+
  ylab("Cockle width (mm)")+
  geom_smooth(method = "lm")+
  facet_wrap(~year)

```

We conclude from this that: \* there were two important outliers in the
dataset that were removed after visual inspection \* the regression
between length and width is about the same for every year, we safely can
use only length as a proxy for the biomass of an individual cockle
